---
title: "EDS 223: Homework 2"
subtitle: ""
author: "Nathalie Bonnet"
date: 2025-10-18
format: html
execute: 
  warning: false
  message: false
  cache: true
---

# Setup Code

1.  Loading necessary libraries

```{r}
#| output: false
# Read in necessary libraries
library(tidyverse)
library(sf)
library(here)
library(tmap)
library(patchwork)
```

2.  Importing data

```{r}
# Load in data
ejscreen <- read_sf(here::here("data", "ejscreen" , "EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"))

redline <- read_sf(here::here("data", "mapping-inequality", "mapping-inequality-los-angeles.json")) %>% 
  st_make_valid() # Validate polygons

gbif <- read_sf(here::here("data", "gbif-birds-LA", "gbif-birds-LA.shp"))
```


Check class of data
```{r}
# Check class of each object
class(ejscreen)

class(redline)

class(gbif)
```


Check crs of each dataframe
```{r}
st_crs(ejscreen)

st_crs(redline)

st_crs(gbif)
```


Compare crs between each 
```{r}
st_crs(ejscreen) == st_crs(redline)

st_crs(ejscreen) == st_crs(gbif) 

st_crs(gbif) == st_crs(redline)
```


Match ejscreen to gbif and redline
```{r}
unique(st_is_valid(ejscreen))
unique(st_is_valid(redline))
unique(st_is_valid(gbif))
```

```{r}
ejscreen <- st_transform(ejscreen, crs = st_crs(redline))
```

Part 1: Legacy of redlining in current environmental (in)justice

```{r}
# Create map of LA by HOLC grade
tm_shape(redline) + # Use redline data
  tm_polygons( # Fill with polygons
    fill.scale = tm_scale(values = c("#7CAF5F", # Colors
                   "#5BA1C6", 
                   "#EDD769", 
                   "#951E00")),
    fill = 'grade', # Fill by 'grade' column
    fill.legend = tm_legend(title = "HOLC grade"), # Add title
    fill_alpha = 0.5 # Make basemap visible under fill layer
  ) +
  tm_basemap("OpenStreetMap") + # Use streetmap data
  tm_title("Historic Home Owners’ Loan Corporation (HOLC) Grades in Los Angeles") + # Set map title
  tm_scalebar(text.size = 0.5) + # Set scalebar size
  tm_compass(position = c("top", "right"), # Put compass on top right corner
             text.size = 0.7) # Compass is larger

```



```{r}
# Make a table representing the percentage of census groups within each grade

# Join datasets using geometries
la_county <- ejscreen %>%
  filter(CNTY_NAME %in% c("Los Angeles County"))


redline_census <- st_join(redline, la_county) %>%
  st_drop_geometry()


# OBJECTID is census block group

grade_census_percent <- redline_census %>%
  group_by(grade) %>%
  summarise(census_percent = (n() / length(redline_census$grade)) * 100)

knitr::kable(
  grade_census_percent,
  col.names = c("HOLC grade", "Percent"),
  caption = "Percent of Los Angeles Census Blocks in each HOLC Grade",
  align = "c"
)


```

Create at least two visualizations summarizing current conditions (from the EJScreen data) within HOLC grades using the mean of the following variables (you may combine variables or create separate plots):
% low income
percentile for Particulate Matter 2.5
percentile for low life expectancy
Use ggplot for your visualizations! You will first need to calculate mean of each variable grouped by HOLC grade.
Write a brief paragraph reflecting on these results
Interpret the patterns you observe in your results
Discuss potential relationships between historical redlining grades and current environmental/socioeconomic conditions

```{r}
means_ejscreen <- redline_census %>% 
  janitor::clean_names() %>% 
  select(grade, lowincpct, pm25, lifeexppct) %>% 
  group_by(grade) %>% 
  summarise(mean_low_inc = (mean(lowincpct)*100), 
            mean_pm25 = mean(pm25), 
            mean_life = (mean(lifeexppct, na.rm = TRUE)*100)) 

lowinc_plot <- ggplot(data = means_ejscreen, aes(x = mean_low_inc, y = grade, fill = grade)) +
  geom_col() +
  scale_fill_manual(values = c("#7CAF5F", "#5BA1C6", "#EDD769", "#951E00", "darkgrey")) +
  theme_minimal() +
  labs(
    title = "Average Percent of Low Income Residents by HOLC Grade in Los Angeles ",
    x = "% low income residents in area",
    y = " ",
    color = "Grade"
  )
  

pm25_plot <- ggplot(data = means_ejscreen, aes(x = mean_pm25, y = grade, fill = grade)) +
  geom_col() +
  scale_fill_manual(values = c("#7CAF5F", "#5BA1C6", "#EDD769", "#951E00", "darkgrey")) +
  theme_minimal() +
  labs(
    title = "Particulate Matter Pollution in Los Angeles by HOLC Grade",
    x = "Average particulate matter air pollution (>2.5 um) in area",
    y = "HOLC grades",
    color = "Grade"
  )

lifeexp_plot <- ggplot(data = means_ejscreen, aes(x = mean_life, y = grade, fill = grade)) +
  geom_col() +
  scale_fill_manual(values = c("#7CAF5F", "#5BA1C6", "#EDD769", "#951E00", "darkgrey")) +
  theme_minimal() +
  labs(
    title = "Percent of Residents with Low Life Expectancy at Birth in Los Angeles by HOLC Grade",
    x = "% of residents in area",
    y = " ",
    color = "Grade"
  )

lowinc_plot + pm25_plot /lifeexp_plot
```


